---
title: "QCcleanup_RecoveryTest_Cyano_pH8"
date: July 26, 2017
output: html_document
---

#This script is used to do QC cleanup, prep data for BMIS, calculate percent recovery from normalized data and plot internal standards.  Required input is QCed Skyline TQS data and the MasterList of HILIC Internal Standards, which contains a column indicating compound type.

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
```

```{r, message=FALSE, warning=FALSE}
datapath <- "~/Desktop/MetabolomicsProjects/SPEvsEtOAcTest/RecoveryTests/Cyano_pH8"
functionpath <- "~/Desktop/MetabolomicsProjects/RScripts"
setwd(datapath)
filename <- "QC_outputCyano_SPErecovery_pH8_pH2_PPL"
data <- read_csv(paste(filename, "csv", sep = "."), skip = 1)
#read_csv("BMIS_Normalized_QC_outputCYANO_TQS_Dissolved_SPEvsEtAc_Aq.csv")



```

#Collect pH8 data
```{r}
data <- data %>% filter(grepl("Poo|pH8|Std|Blk", Replicate.Name))
```


#Load Internal Standard List
```{r}
setwd("~/Desktop/MetabolomicsProjects/SPEvsEtOAcTest/RecoveryTests/Cyano_pH8")
MasterList <- read.csv("CYANO_MasterList_March2017.csv")

MasterList <- MasterList %>%
  select(Group, Compound.Name)

Unique.MasterList <- unique(MasterList)

IS.list <- MasterList %>% filter(Group == "Internal Std") %>% 
     select(`Compound.Name`) %>% 
     unique()
IS.list <- IS.list[[1]]
IS.list <- as.character(IS.list)


```
#QC cleanup
```{r}
data_new <- data %>%
  separate(Replicate.Name,c("runDate","type","SampID","replicate"),"_", remove=FALSE)

#A few bad compounds made it through QC
# data_new <- data_new %>%
#   filter(!Compound.Name == "Phosphocholine",
#          !Compound.Name == "Tryptophol",
#          !Compound.Name == "7-dehydrocholesterol",
#          !Compound.Name == "Vitamin D2",
#          !Compound.Name == "Vitamin C",
#          !Compound.Name == "Methylthioadenosine",
#          !Compound.Name == "2, 4 decadienal",
#          !Compound.Name == "DHA")

Unique.names <-unique(data_new$Compound.Name)

Check.data = data.frame(matrix(vector(), 0, 5,
                               dimnames=list(c(), c("Compound.Name", "is.NA", "NumNA", "per.NA", "ALLNA" ))),
                        stringsAsFactors=F)

Names <- colnames(data_new)

temp.good = data.frame(matrix(vector(), 0, length(Names),
                                        dimnames=list(c(), c(Names))),
                                 stringsAsFactors=F)

temp.good.pooled <- temp.good

data_filt <- data_new %>% filter(type == "Smp")

#Check to see if any compounds are not detected in any SAMPLES

for (i in 1:length(Unique.names))
{Compound_set <- data_filt %>% filter(Compound.Name == Unique.names[i])

Check.data[i,1] <- as.character(Unique.names[i])

if (anyNA(Compound_set$Area)==TRUE)
{
  Check.data[i,2] <-TRUE
  Check.data[i,3] <-sum(is.na(Compound_set$Area))
  Check.data[i,4] <-sum(is.na(Compound_set$Area))/length(Compound_set$Compound.Name)
  {
    if (all(is.na(Compound_set$Area))==TRUE)
      Check.data[i,5] <-TRUE
  }
}

else  {
  Check.data[i,2] <-FALSE
  
}

}

#Make a list of compounds we detected

Check.data_trim <- Check.data %>% filter(is.na(ALLNA))

#OR
#Check.data_trim <- Check.data %>% 
# filter(per.NA < 0.5 | is.NA == "FALSE")
# 
data_blanks <- data_new %>% filter(type == "Blk")
data_standards <- data_new %>% filter(type == "Std")
data_pooled <- data_new %>% filter(type == "Poo")

#replace NAs in SAMPLES only (does not back fill pooled, standards, or blanks)

for (i in 1:length(Check.data_trim$Compound.Name)) {
  
  NA_value <- data_blanks %>% 
    filter(Compound.Name == Check.data_trim[i,1]) %>% 
    select(Area) 
  
  temp <- data_filt %>% 
    filter(Compound.Name == Check.data_trim[i,1]) %>% 
    replace_na(list(Area=NA_value[1,1]))
  

  
  temp.good<- rbind(temp.good, temp)

}



remove_from_standards <- Check.data %>% filter(!is.na(ALLNA))

data_standards <- data_standards %>% 
  filter(!Compound.Name %in% remove_from_standards[,1])

data_blanks <- data_blanks %>% 
  filter(!Compound.Name %in% remove_from_standards[,1])

data_pooled <- data_pooled %>% 
  filter(!Compound.Name %in% remove_from_standards[,1])

#Wouldn't normally need to do this next step, but a few wierd things got through the QC and we need to backfill missing Area in a few pooled 
# data_pooled <- data_pooled %>% 
#     mutate(Area = ifelse(is.na(data_pooled$Area), data_pooled$rawArea, data_pooled$Area))



drop <- c("runDate","type","SampID","replicate")

data_blankreplaced <- temp.good %>% 
  rbind(data_blanks, data_standards, data_pooled) %>% 
  select(-one_of(drop)) %>% 
  mutate(Area = as.numeric(Area))




setwd(datapath)
write.csv(data_blankreplaced, file = paste(filename, "_NAsReplaced", ".csv", sep = "" ), row.names = FALSE)
#
#data_blankreplaced <- temp.good %>% rbind(data_blanks, data_standards) %>% mutate(Area = as.integer(Area))
```

#DO B-MIS-----

```{r}
setwd(datapath)
filename <- "QC_outputCyano_SPErecovery_pH8_PPL"
#load old data_blankreplaced
data_blankreplaced <- read.csv(file = paste(filename, "_NAsReplaced", ".csv", sep = "" ), row.names=1)

#load BMIS normalized data
#
data_normalized <- read.csv(file = paste("New_Normalized_data_", filename, "_NAsReplaced", ".csv", sep = "" ))
```

#Plot selected compounds
```{r}
data_norm_subset <- data_normalized %>% 
  filter(Compound.Name == "GA3" |
           Compound.Name == "2iP" |
           Compound.Name == "Hydroxo B12"|
           Compound.Name == "Cyano B12" |
           Compound.Name =="GA4" |
           Compound.Name == "Sulfurol" |
           Compound.Name == "Acetovanillone") %>% 
  filter(!grepl("Std", Replicate.Name))

p <- ggplot(data_norm_subset, aes(x=Replicate.Name, y=PooPlusModel)) + geom_bar(stat="identity") + facet_wrap( ~Compound.Name, scales="free_y")+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5), legend.position = "top")  + ggtitle("Select Compounds Normalized Areas Cyano pH8")

q <- ggplot(data_norm_subset, aes(x=Replicate.Name, y=Area)) + geom_bar(stat="identity") + facet_wrap( ~Compound.Name, scales="free_y")+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5), legend.position = "top")  + ggtitle("Select Compounds Raw Areas Cyano pH8")

```

```{r}
data_GA3_allNormers <- data_normalized %>% 
  filter(Compound.Name == "GA3") %>% 
  filter(!grepl("Std", Replicate.Name)) %>%
  select(Replicate.Name, Area, inj.vol.Norm.Area, d4.Tryptamine.Norm.Area, d.IAA.Norm.Area, D.Phenylalanine.Norm.Area, D.Tryptophan.Norm.Area, Vitamin.B1_IS.Norm.Area, Vitamin.B2_IS.Norm.Area, D8.Arachidonic.Acid.Norm.Area, D5.EPA.Norm.Area, D5.DHA.Norm.Area, D6.Tocopherol.Norm.Area, D3.Pyridoxal.Norm.Area) %>% 
  gather(Area, inj.vol.Norm.Area, d4.Tryptamine.Norm.Area, d.IAA.Norm.Area, D.Phenylalanine.Norm.Area, D.Tryptophan.Norm.Area, Vitamin.B1_IS.Norm.Area, Vitamin.B2_IS.Norm.Area, D8.Arachidonic.Acid.Norm.Area, D5.EPA.Norm.Area, D5.DHA.Norm.Area, D6.Tocopherol.Norm.Area, D3.Pyridoxal.Norm.Area, key = "Normer", value = "Area")

r <- ggplot(data_GA3_allNormers, aes(x=Replicate.Name, y=Area)) + geom_bar(stat="identity") + facet_wrap( ~Normer, scales="free_y")+theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust = 0.5), legend.position = "top")  + ggtitle("All IS Normalizations for GA3, Cyano pH8")
```


#Calculate Recovery

```{r}
#Substract "no spike" avg value from treatment values

#Create a data.frame of average background values seen in the nospike samples
data_backgroundvalues <- data_normalized %>%
  filter(grepl('nospike', Replicate.Name))%>%
  mutate(Compound.Name = as.factor(Compound.Name)) %>%
  group_by(Compound.Name) %>%
    dplyr::summarise(mean.background = mean(PooPlusModel))

data_backgroundcorrected <- data_normalized %>%
   filter(!grepl('nospike', Replicate.Name))

Unique.names <-unique(data_blankreplaced$Compound.Name)

Names <- colnames(data_backgroundcorrected)

temp.good = data.frame(matrix(vector(), 0, length(Names),
                                        dimnames=list(c(), c(Names))),
                                 stringsAsFactors=F)

#calculate background-corrected areas (area.corrected) by subracting nospike values (PooPlusModel) from spikebefore and spikeafter samples (PooPlusModel)
for (i in 1:length(Unique.names)) {
  
  data_backgroundvalue <- data_backgroundvalues %>%
    filter(Compound.Name == Unique.names[i]) 
  
  data_backgroundvalue <- as.data.frame(data_backgroundvalue)
   data_backgroundcorrected <- as.data.frame(data_backgroundcorrected)
  
  Compound_set <- data_backgroundcorrected %>% 
    filter(grepl('Smp', Replicate.Name)) %>%
    filter(!grepl("spikeafter_2", Replicate.Name)) %>%
    filter(!grepl("spikeafter_3", Replicate.Name)) %>%
    filter(Compound.Name == Unique.names[i]) %>%
    mutate(area.corrected = (PooPlusModel - data_backgroundvalue[1,2]))

  
  temp.good<- rbind(temp.good, Compound_set)

}
#temp.good_backup <- temp.good 

#if area.corrected is less than zero, replace with zero
temp.good$area.corrected <- ifelse(grepl("spikebefore", temp.good$Replicate.Name) & temp.good$area.corrected < 0, 0, temp.good$area.corrected)



#Calculate percent reovery in 'spikebefore' samples relative to 'spikeafter' samples
#'spikeafter' is multiplied by 2 because it was prepared to a final 2 uM concentration, rather than 4 uM. spikebefore would be 4 uM if 100% recovery
Recovery <- temp.good %>%
  #separate(Replicate.Name,c("runDate","type","SampID","replicate"),"_", remove=FALSE) %>%
  group_by(Compound.Name, SampID) %>%
  dplyr::summarise(mean.area.corrected = mean(area.corrected)) %>%
  mutate(SampID = gsub('SPEpH8-', '', SampID)) %>%
  spread(key = SampID, value=mean.area.corrected) %>%
  mutate(percent.recovery = (spikebefore/(spikeafter1*2))*100) %>%
  mutate(percent.recovery = round(percent.recovery)) %>%
  arrange(desc(percent.recovery)) %>%
  left_join(Unique.MasterList, by="Compound.Name")

Recovery_trim <- Recovery %>%
  select(Compound.Name, percent.recovery)


setwd(datapath)
  write.csv(Recovery, "RecoveryValues_postBMIS_pH8_Cyano_NOVB1.csv")
```

#Plot internal standards

```{r}
internal.standards.list <- Unique.MasterList %>%
  filter(Group == "Internal Std")

internal.standards <- data_normalized %>%
  # mutate(Replicate.Name = gsub('spikeafter_1', 'spikeafter-1_', Replicate.Name)) %>%
  # mutate(Replicate.Name = gsub('spikeafter_2', 'spikeafter-2_', Replicate.Name)) %>%
  # mutate(Replicate.Name = gsub('spikeafter_3', 'spikeafter-3_', Replicate.Name)) %>%
  # separate(Replicate.Name,c("runDate","type","SampID","replicate"),"_", remove=FALSE) %>%
  #mutate(SampID = gsub('SPEpH5-', '', SampID)) %>%
  filter(grepl('Smp', Replicate.Name)) %>%
  filter(Compound.Name %in% internal.standards.list$Compound.Name) %>%
  group_by (Compound.Name, SampID) %>%
  summarise(mean.Area = mean(PooPlusModel), sd = sd(PooPlusModel)) 


 p <- ggplot(internal.standards, aes(y=mean.Area, x=SampID)) + 
      geom_bar(stat="identity", fill="white", colour="black", width = 0.5, position = "dodge") +
      geom_errorbar(aes(ymin=mean.Area-sd, ymax=mean.Area+sd), width=0.2) +
      #labs(y="Area", x="SampID") +
      scale_y_continuous(expand = c(0,0))+
      theme(axis.text.x=element_text(angle=50, size=8, vjust=0.5)) +
      facet_wrap(~Compound.Name, scales= "free", ncol=3)

 
 print(p)

```
#Which compounds are present in background seawawater?

```{r}
data_nospike <- data %>%
  filter(grepl("nospike", Replicate.Name)) %>%
  filter(!Compound.Name %in% IS.list)

Unique.names <-unique(data_nospike$Compound.Name)

Check.data = data.frame(matrix(vector(), 0, 2,
                               dimnames=list(c(), c("Compound.Name", "NumNA"))),
                        stringsAsFactors=F)


for (i in 1:length(Unique.names))
{Compound_set <- data_nospike %>% filter(Compound.Name == Unique.names[i])

Check.data[i,1] <- as.character(Unique.names[i])

Check.data[i,2] <-sum(is.na(Compound_set$Area))
}

keep <- Check.data %>%
  filter(NumNA < 2)

data_nospike <- data_nospike %>%
  filter(Compound.Name %in% keep$Compound.Name)

```


#Build calibration curve for compounds present in seawater background
```{r}
data_calibration <- data_normalized %>%
  filter(grepl("spikeafter", Replicate.Name)) %>%
  filter(Compound.Name %in% keep$Compound.Name) %>%
  select(Replicate.Name, Compound.Name, PooPlusModel.IS, PooPlusModel) %>%
  mutate(concentration = ifelse(grepl("spikeafter1", Replicate.Name), "2",
                          ifelse(grepl("spikeafter2", Replicate.Name), "1",
                           ifelse(grepl("spikeafter3", Replicate.Name), "0.5",""))))%>%
  select(-Replicate.Name) %>%
  group_by(Compound.Name, concentration) %>%
  dplyr::summarise(mean.PooPlusModel = mean(PooPlusModel)) %>%
  spread(key=concentration, value=mean.PooPlusModel) %>%
  mutate(intercept = NA) %>%
  mutate(slope = NA) %>%
  mutate(r.squared = NA) %>% 
  mutate(slope_0 = NA) %>%
  mutate(r.squared_0 = NA)

unique.compounds <- as.character(unique(data_calibration$Compound.Name))

data_calibration2 <- data_normalized %>%
  filter(grepl("spikeafter", Replicate.Name)) %>%
  filter(Compound.Name %in% keep$Compound.Name) %>%
  select(Replicate.Name, Compound.Name, PooPlusModel.IS, PooPlusModel) %>%
  mutate(concentration = ifelse(grepl("spikeafter1", Replicate.Name), "2",
                          ifelse(grepl("spikeafter2", Replicate.Name), "1",
                           ifelse(grepl("spikeafter3", Replicate.Name), "0.5",""))))%>%
  mutate(concentration = as.numeric(concentration))

for (i in 1:length(unique.compounds)) {
  
  Compound_set <- data_calibration2 %>% filter(Compound.Name == unique.compounds[i])
  
  y <- Compound_set$PooPlusModel
  x <- Compound_set$concentration
  
  model <- lm(y ~ x, Compound_set)
  model_0 <- lm(y ~ x -1, Compound_set)
  
  data_calibration[data_calibration$Compound.Name == unique.compounds[i], "slope"] <- model$coefficients[2]
  
   data_calibration[data_calibration$Compound.Name == unique.compounds[i], "intercept"] <- model$coefficients[1]
   
   data_calibration[data_calibration$Compound.Name == unique.compounds[i], "r.squared"] <- summary(model)$r.squared
   
   data_calibration[data_calibration$Compound.Name == unique.compounds[i], "r.squared_0"] <- summary(model_0)$r.squared
  
   data_calibration[data_calibration$Compound.Name == unique.compounds[i], "slope_0"] <- model_0$coefficients[1]
  
}

data_calibration_trim <- data_calibration %>%
  select(Compound.Name, intercept, slope, r.squared, slope_0, r.squared_0)
```

#plot calibration curves

```{r}
setwd(functionpath)
source("FacetGraphs_RecoveryTest.R")

Compound.Names <- as.character(unique(data_calibration2$Compound.Name))
length(Compound.Names)
Compound.Names_1 <-Compound.Names[1:12]
Compound.Names_2 <-Compound.Names[13:24]
#Compound.Names_3 <-Compound.Names[25:36]

data_calibration2_1 <- data_calibration2 %>% filter(Compound.Name %in% Compound.Names_1)
data_calibration2_2 <- data_calibration2 %>% filter(Compound.Name %in% Compound.Names_2)
#data_calibration2_3 <- data_calibration2 %>% filter(Compound.Name %in% Compound.Names_3)

a <- FacetGraphs(data_calibration2_1)

b <- FacetGraphs(data_calibration2_2)
#c <- FacetGraphs(data_calibration2_3)

```



#Calculate concentrations present in background seawater

```{r}
low.concentration <- data_calibration %>%
  select(Compound.Name, `0.5`)

blank.value <- data_normalized %>%
  filter(grepl("Blk", Replicate.Name)) %>% 
  select(Compound.Name, PooPlusModel) %>% 
  dplyr::rename(Blank.Area = PooPlusModel)
  

data_nospike_conc <- data_normalized %>%
  filter(Compound.Name %in% keep$Compound.Name) %>%
  filter(grepl("nospike", Replicate.Name)) %>% 
  select(Replicate.Name, Compound.Name, PooPlusModel.IS, PooPlusModel) %>% 
  left_join(data_calibration_trim, "Compound.Name") %>%
  left_join(Recovery_trim, "Compound.Name") %>%
  mutate(raw.concentration = (PooPlusModel - intercept)/slope)%>%
   mutate(raw.concentration_0 = PooPlusModel/slope_0)%>%
  mutate(adjusted.concentration = raw.concentration_0/(percent.recovery/100)) %>%
  mutate(raw.concentration = as.numeric(format(raw.concentration, digits = 2, scientific=FALSE))) %>% 
  mutate(raw.concentration_0 = as.numeric(format(raw.concentration_0, digits = 2, scientific=FALSE))) %>% 
   left_join(Unique.MasterList, by="Compound.Name") %>% 
  left_join(low.concentration, by="Compound.Name") %>% 
  left_join(blank.value, by="Compound.Name") %>% 
  mutate(PooPlusModel =format(PooPlusModel, digits = 2, scientific =TRUE)) %>% 
  mutate(`0.5` =format(`0.5`, digits = 2, scientific =TRUE)) %>% 
  mutate(Blank.Area =format(Blank.Area, digits = 2, scientific =TRUE)) %>% 
  select(Replicate.Name, Compound.Name, percent.recovery, raw.concentration_0, adjusted.concentration, r.squared_0, Group, PooPlusModel, `0.5`, Blank.Area, raw.concentration, r.squared, everything()) %>% 
  filter(percent.recovery > 9) %>% 
  group_by(Compound.Name, percent.recovery) %>% 
  dplyr::summarise(average.adj.conc = mean(adjusted.concentration), 
                   sd.adj.conc = sd(adjusted.concentration),
                   average.raw.conc = mean(raw.concentration_0),
                   sd.raw.conc = sd(raw.concentration_0)) 

setwd(datapath)
write_csv(data_nospike_conc, "BackgroundConcentrations_SPE_pH8_Cyano.csv")
#build an adjusted.concentration
#If no spike mean(raw area) is not twice background (or other multiplier), discard
#If percent.recovery is less than 10, remove compound
#If raw.concentration is negative, replace with zero
#  mutate(adjusted.concentration = raw.concentration/(percent.recovery/100))
```


